Sparse spectral and $p$-finite element methods for partial differential equations on spherical caps

### 2020-08-20 papers/spherical-caps/spherical-caps.tex
Draft of paper, with pdf figs included. Should be able to compile.

As in the work in two dimensions on the show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that are defined in such a way that mirrors the approach on quadratic curves \cite{olver2020orthogonal}. On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k - \tilde k_i} \: y^{\tilde k_i} \: z^{n-k} + (\hbox{lower order terms})
\end{align*} \bstodo{Ordering correct?}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - \tilde j_i} \: y^{\tilde j_i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$. In this work, as for the disk-slice case in 2D \cite{snowball2019sparse}, we use a more general integration-by-parts argument to deduce the sparsity structure. The 3D OPs defined here involve the same non-classical (in fact, semi-classical) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with one-dimensional orthogonal polynomials we can construct discretizations of general partial differential operators of size $p(p-1)/2 \times p(p-1)/2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This compares favourably to $O(p^6)$ operations if one proceeds na\"ively. Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}. \bstodo{reword to avoid plagiarising self}

@@

\begin{definition}
Define $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ as the 2D OPs on $(x,y)\in\R$ s.t. $x^2 + y^2 = 1$; that is, if we define $\theta$ by $x = \cos(\theta)$, $y = \sin(\theta)$ for $\xvec := (x, y)$ where $||\xvec|| = 1$,
\begin{align}
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) \equiv \ch_{0,0}(\theta) &:= \frac{\sqrt{2}}{2} =: \ch_{0} \nonumber \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) \equiv \ch_{k,0}(\theta) &:= \cos(k \theta), \quad k = 1,2,3,\dots\label{eqn:circularharmonics} \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) \equiv \ch_{k,0}(\theta)& := \sin(k \theta), \quad k = 1,2,3,\dots \nonumber
\end{align}
orthonormal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_{||\xvec|| = 1} p(\xvec) \: q(\xvec) \: \D \sigma(\xvec) = \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta. \label{eqn:ipcircularharmonics}
\end{align}
\end{definition}
Of course, these are simply Chebyshev polynomials $\ch_{k,0}(x, y) = T_k(x)$ and $\ch_{k,1}(x, y) = y \: U_{k-1}(x)$ for $k > 0$, where we define $\ch_0$ so as to ensure orthonormality.

\begin{proposition}\label{prop:construction}
Let $w : (\alpha,\beta) \: \to \R$ be a weight function. $\forall$, $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$
\begin{align*}
	\scopnki(x,y,z) \equiv \scopnki(\xvec,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr(\frac{\xvec}{\rho(z)})
\end{align*}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\begin{align*}
	\ip< p, \: q > &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: w(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: w(z) \: \D z.
\end{align*}
on $\Omega$. 
\end{proposition}

@@

For the spherical cap, we can use Proposition \ref{prop:construction} to create our one-parameter family of OPs. We first introduce notation for our family of non-classical univariate OPs that will be used as the $r_n$ polynomials above.
\begin{definition}[\cite{snowball2019sparse}]\label{def:OPconstruction}
Let $\genjacw^{(a,b)}(x)$ be a weight function on the interval $(\alpha, \beta)$ given by:
\begin{align*}
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{b} \: \rho(x)^{b}
\end{align*}
and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipgenjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b)} := \int_\alpha^\beta \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipnormalisation}
\end{align}
is a normalising constant.
Denote the two-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}).
\end{definition}

@@

By construction, $\{\scopnkia\}$ are orthogonal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\scopa} &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: \genjacw^{(a,0)}(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: \genjacw^{(a,0)}(z) \: \D z,
\end{align}
with
\begin{align}
	\norm{\scopnkia}_{\scopa}^2 &:= \ip< \scopnkia, \: \scopnkia >_{\scopa} \nonumber \\
	&= \pi \: \normgenjac^{(a, 2k)} \label{eqn:normscop}
\end{align}
We note that the weight $\genjacw^{(a,b)}(z)$ has been used in the construction of 2D orthogonal polynomials on disk-slices and trapeziums \cite{snowball2019sparse}, where a method for obtaining recurrence coefficients and evaluating integrals was established (the weight is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).

For clarity we state here that throughout we will regularly use coordinate transformations, mainly in any integrals over the surface $\Omega$, between the cartesian $(x,y,z)$ and the spherical coordinates $(\phi, \theta)$ -- that is, for any point $(x,y,z) \in \Omega$, let $(\phi, \theta)$ denote the angles such that:
\begin{align*}
	z &= \cos(\phi) \\
	x &= \sin(\phi) \: \cos(\theta) = \rho(z) \: \cos(\theta) \\
	y &= \sin(\phi) \: \sin(\theta) = \rho(z) \: \sin(\theta).
\end{align*}

@@

It follows that we can apply $\Dnt$ in $O(n)$ complexity, and thereby calculate $\bigscopta_{0}(x,y)$  through $\bigscopta_{n}(x,y)$ in optimal $O(n^2)$ complexity. \bstodo{correct for this case}

\begin{definition}\label{def:clenshawmats}
The Clenshaw matrices associated with the OPs $\{\scopnkia\}$ given by the matrices $A_n, B_n, C_n, \Dnt$ for $n = 0,1,2,\dots$ defined above.
\end{definition}

@@

	\medskip
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-helmholtz}
		\centering
		\caption{A variable coefficient Helmholtz operator}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-biharmonic}
		\centering
		\caption{The Biharmonic operator $B_W^{(2)}$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity. For (c), the the weighted variable coefficient Helmholtz operator is $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{(x_0^2 + z_0^2)}$ and $k = 200$.}
	\label{fig:images}
\end{figure}

General linear partial differential operators with polynomial variable coefficients can be constructed by composing the sparse representations for partial derivatives, conversion between bases, and Jacobi operators. As a canonical example, we can obtain the matrix operator for the $\rho^2$-factored surface Laplacian $\rho(z)^2 \: \nabla^2_s$, that will take us from coefficients for expansion in the weighted space $\bigW_N^{(1)}(x,y,z) = \genjacw^{(1,0)}(z) \: \bigscopN^{(1)}(x,y,z)$ to coefficients in the non-weighted space $\bigscopN^{(1)}(x,y,z)$. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$, similar to how the Dirichlet zero boundary conditions would be imposed for the operator $\Delta^{(1)}_W$ in Definition \ref{def:differentialoperators}. The matrix operator for this $\rho^2$-factored Laplacian we denote $\mathcal{L}_W^{(1)}$ acting on the coefficients vector is then given by
\begin{align*}
    \mathcal{L}_W^{(1)} := D_\phi^{(0)} \: W_\phi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2.
\end{align*}
Importantly, this operator will have banded-block-banded structure, and hence will be sparse, as seen in Figure \ref{fig:sparsity}. \bstodo{Add this fig}

@@

For a function $f$, with coefficients vector $\bm{f}$ for expansion in the $\{\scopnki\}$ basis as determined via the method in Section \ref{subsection:expandingfunctions} up to order $N$, we can use the Clenshaw algorithm to evaluate the function at a point $(x,y,z) \in \Omega$ as follows. Let $A_n, B_n, \Dnt, C_n$ be the Clenshaw matrices from Definition \ref{def:clenshawmats}, and define the rearranged coefficients vector $\tilde{\bm{f}}$ via
\begin{align*}
	\tilde{\bm{f}}_n &:= 
		\begin{pmatrix}
			f_{n,0,0} \\
			f_{n,1,0} \\
			f_{n,1,1} \\
			\vdots \\
			f_{n,n,0} \\
			f_{n,n,1}
		\end{pmatrix} \in \R^{2(N+1)} \quad \text{for } n = 1,2,\dots,N, \quad
	\tilde{\bm{f}}_0 = f_{0,0,0} \in \R, \\
	\tilde{\bm{f}} &:= 
		\begin{pmatrix}
			\tilde{\bm{f}}_0 \\
			\vdots \\
			\tilde{\bm{f}}_N
		\end{pmatrix} \in \R^{(N+1)^2}.
\end{align*}
The Clenshaw algorithm is then \cite{}\bstodo{Reference for Clenshaw in >1 dimensions???} 
\begin{align*}
	\quad &\text{1) } \text{Set } \bm{\xi}_{N+2} = \bold{0}, \: \bm{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \bm{\xi}_{n}^T = {\tilde{\bm{f}}_n}^T - \bm{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \bm{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \bm{\xi}_{0} \: \bigscopta_{0}  = \xi_{0} \: \scopa_0
\end{align*}

@@

The proofs of Theorem \ref{theorem:sparsityofdifferentialoperators} and Lemma \ref{lemma:sparsityofparametertransformationoperators} provide a way to calculate the non-zero entries of the operator matrices given in Definition \ref{def:differentialoperators} and Definition \ref{def:parametertransformationoperators}. We can simply use quadrature to calculate the 1D inner products, which has a complexity of $\bigO(N^3)$. This proves much cheaper computationally than using the 3D quadrature rule to calculate the 3D inner products, which has a complexity of $\bigO(N^4)$ \bstodo{correct the O(nd) complexity}.

@@

% FIGURES
\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.17]{poisson-u=wyexpx-N=60}
	\centering
	%\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.3]{solutionblocknorms-poisson-N=100}
        	%\label{fig:solutionblocknorms-poisson}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y,z) = - 2 e^x y z (2+x) + \genjacw^{(1,0)}(z) e^x (y^3 + z^2 y - 4xy - 2y)$. Right: The norms of each block of the computed solution of the Poisson equation with the given right hand side functions (here, $\bf{p}$ is simply a point on the spherical cap surface away from the pole and boundary). This demonstrates spectral convergence observed.}
	\centering
	\label{fig:poisson}
\end{figure}
%
%\begin{figure}[t]
%	\begin{subfigure}{0.3\textwidth}
%	\includegraphics[scale=0.3]{solution-exact-poisson-diskslice-alpha=0p2-beta=0p8-u=y3expx}
%	\centering
%	\end{subfigure}
%	\begin{subfigure}{0.3\textwidth}
%	\centering
%	\includegraphics[scale=0.3]{solution-exact-poisson-diskslice-alpha=0p2-beta=0p8-u=y3expx-actual}
%	\centering
%	\end{subfigure}
%	\begin{subfigure}{0.3\textwidth}
%	\includegraphics[scale=0.3]{solution-exact-poisson-diskslice-alpha=0p2-beta=0p8-u=y3expx-compare}
%	\centering
%	\end{subfigure}
%	\centering
%	\caption{The computed solution to $\Delta u = f$ with zero boundary conditions compared with the exact solution $u(x,y) = \Wiii (x,y) y^3 \exp(x)$. Left: Computed. Centre: Exact. Right: Plot of the error (colourbar is shown to demonstrate magnitude of the error is of the order $10^{-17}$)}
%	\centering
%	\label{fig:poissonexact}
%\end{figure}

\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.17]{helmholtz-f=wyexpx-N=60-n=300}
	\centering
	%\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.3]{solutionblocknorms-helmholtz-N=100}
        	%\label{fig:solutionblocknorms-poisson}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y,z) = y e^x \genjacw^{(1,0)}(z)$, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{(x_0^2 + z_0^2)}$ and $k = 100$. Right: The norms of each block of the computed solution of the Helmholtz equation with the given right hand side functions, with $k=20$ and the same function $v$ (again, $\bf{p}$ is simply a point on the spherical cap surface away from the pole and boundary). This demonstrates spectral convergence observed.}
	\centering
	\label{fig:helmholtz}
\end{figure}

>>>>

As in the work in two dimensions on the show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that are defined in such a way that mirrors the approach on quadratic curves \cite{olver2020orthogonal}. On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k - \tilde k_i} \: y^{\tilde k_i} \: z^{n-k} + (\hbox{lower order terms})
\end{align*} \bstodoinline{How can I write down what the ordering is? Is this necessary?}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - \tilde j_i} \: y^{\tilde j_i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$. In this work, as for the disk-slice case in 2D \cite{snowball2019sparse}, we use a more general integration-by-parts argument to deduce the sparsity structure. The three-dimensional orthogonal polynomials defined here involve the same non-classical (in fact, semi-classical) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with these 1D OPs we can construct discretizations of general partial differential operators of size $(p+1)^2 \times (p+1)^2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This clearly compares favourably to proceeding in a na\"ive approach where one would require $O(p^6)$ operations \bstodoinline{Presumably this is correct to say?}. Furthermore, we can use this framework to derive sparse $p$-finite element methods, which we introduce as an Appendix here. The methods are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

@@

\begin{definition}
Define the unit circle $\omega := \{ \mathbf{x} = (x,y) \in \R^2 \: | \: x^2 + y^2 = 1\}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos(\theta)$, $y = \sin(\theta)$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\begin{align}
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \nonumber \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) \ &:= T_k(x) = \cos(k \theta) =: \ch_{k,0}(\theta), \quad k = 1,2,3,\dots\label{eqn:chebyshev} \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin(k \theta) =: \ch_{k,1}(\theta), \quad k = 1,2,3,\dots \nonumber
\end{align}
where $\ch_{0} := \frac{\sqrt{2}}{2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_{||\xvec|| = 1} p(\xvec) \: q(\xvec) \: \D \sigma(\xvec) = \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta. \label{eqn:ipcircularharmonics}
\end{align}
\end{definition}
Note that we have defined $\ch_0$ so as to ensure orthonormality.

\begin{proposition}\label{prop:construction}
Let $w : (\alpha,\beta) \: \to \R$ be a weight function. $\forall$, $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$
\begin{align*}
	\scopnki(x,y,z) \equiv \scopnki(\xvec,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr(\frac{\xvec}{\rho(z)})
\end{align*}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\begin{align*}
	\ip< p, \: q > &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: w(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: w(z) \: \D z
\end{align*}
on $\Omega$. 
\end{proposition}

@@

For the spherical cap, we can use Proposition \ref{prop:construction} to create our one-parameter family of OPs. We first introduce notation for our family of non-classical univariate OPs that will be used as the $r_n$ polynomials above.
\begin{definition}[\cite{snowball2019sparse}]\label{def:OPconstruction}
Let $\genjacw^{(a,b)}(x)$ be a weight function on the interval $(\alpha, 1)$ given by:
\begin{align*}
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{a} \: \rho(x)^{b}
\end{align*}
and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipgenjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b)} := \int_\alpha^1 \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipnormalisation}
\end{align}
is a normalising constant.
Denote the two-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}).
\end{definition}

@@

By construction, $\{\scopnkia\}$ are orthogonal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\scopa} &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: \genjacw^{(a,0)}(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: \genjacw^{(a,0)}(z) \: \D z,
\end{align}
with
\begin{align}
	\norm{\scopnkia}_{\scopa}^2 &:= \ip< \scopnkia, \: \scopnkia >_{\scopa} = \pi \: \normgenjac^{(a, 2k)} \label{eqn:normscop}
\end{align}
We note that the weight $\genjacw^{(a,b)}(z)$ has been used in the construction of 2D orthogonal polynomials on disk-slices and trapeziums \cite{snowball2019sparse}, where a method for obtaining recurrence coefficients and evaluating integrals was established (the weight is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).

For clarity we state here that throughout we will regularly use coordinate transformations, mainly in any integrals over the surface $\Omega$, between the cartesian $(x,y,z)$ and the spherical coordinates $(\phi, \theta)$ -- that is, for any point $(x,y,z) \in \Omega$, let $(\phi, \theta)$ denote the angles such that:
\begin{align*}
	z &= \cos(\phi) \\
	x &= \sin(\phi) \cos(\theta) = \rho(z) \cos(\theta) \\
	y &= \sin(\phi) \sin(\theta) = \rho(z) \sin(\theta).
\end{align*}

@@

It follows that we can apply $\Dnt$ in $O(n)$ complexity, and thereby calculate $\bigscopta_{0}(x,y,z)$  through $\bigscopta_{n}(x,y,z)$ in optimal $O(n^2)$ complexity. \bstodoinline{I think this is correct for this case}

\begin{definition}\label{def:clenshawmats}
The Clenshaw matrices associated with the OPs $\{\scopnkia\}$ are given by the matrices $A_n, B_n, C_n, \Dnt$ for $n = 0,1,2,\dots$ defined above.
\end{definition}

@@

	\medskip
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-helmholtz}
		\centering
		\caption{A variable coefficient Helmholtz operator}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[scale=0.3]{sparsity-of-biharmonic}
		\centering
		\caption{The Biharmonic operator $B_W^{(2)}$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity. For (c), the the weighted variable coefficient Helmholtz operator is $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{(x_0^2 + z_0^2)}$ and $k = 200$.}
	\label{fig:sparsity}
\end{figure}

General linear partial differential operators with polynomial variable coefficients can be constructed by composing the sparse representations for partial derivatives, conversion between bases, and Jacobi operators. As a canonical example, we can obtain the matrix operator for the $\rho^2$-factored surface Laplacian $\rho(z)^2 \: \nabla^2_s$, that will take us from coefficients for expansion in the weighted space $\bigW_N^{(1)}(x,y,z) = \genjacw^{(1,0)}(z) \: \bigscopN^{(1)}(x,y,z)$ to coefficients in the non-weighted space $\bigscopN^{(1)}(x,y,z)$. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$, similar to how the Dirichlet zero boundary conditions would be imposed for the operator $\Delta^{(1)}_W$ in Definition \ref{def:differentialoperators}. The matrix operator for this $\rho^2$-factored Laplacian we denote $\mathcal{L}_W^{(1)}$ acting on the coefficients vector is then given by
\begin{align*}
    \mathcal{L}_W^{(1)} := D_\phi^{(0)} \: W_\phi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2.
\end{align*}
Importantly, this operator will have banded-block-banded structure, and hence will be sparse, as seen in Figure \ref{fig:sparsity}.

@@

For a function $f$, with coefficients vector $\bm{f}$ for expansion in the $\{\scopnki\}$ basis as determined via the method in Section \ref{subsection:expandingfunctions} up to order $N$, we can use the Clenshaw algorithm to evaluate the function at a point $(x,y,z) \in \Omega$ as follows. Let $A_n, B_n, \Dnt, C_n$ be the Clenshaw matrices from Definition \ref{def:clenshawmats}, and define the rearranged coefficients vector $\tilde{\bm{f}}$ via
\begin{align*}
	\tilde{\bm{f}}_n &:= 
		\begin{pmatrix}
			f_{n,0,0} \\
			f_{n,1,0} \\
			f_{n,1,1} \\
			\vdots \\
			f_{n,n,0} \\
			f_{n,n,1}
		\end{pmatrix} \in \R^{2(N+1)} \quad \text{for } n = 1,2,\dots,N, \quad
	\tilde{\bm{f}}_0 = f_{0,0,0} \in \R, \\
	\tilde{\bm{f}} &:= 
		\begin{pmatrix}
			\tilde{\bm{f}}_0 \\
			\vdots \\
			\tilde{\bm{f}}_N
		\end{pmatrix} \in \R^{(N+1)^2}.
\end{align*}
The Clenshaw algorithm is then \cite{}\bstodoinline{Reference for Clenshaw in >1 dimensions???} 
\begin{align*}
	\quad &\text{1) } \text{Set } \bm{\xi}_{N+2} = \bold{0}, \: \bm{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \bm{\xi}_{n}^T = {\tilde{\bm{f}}_n}^T - \bm{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \bm{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \bm{\xi}_{0} \: \bigscopta_{0}  = \xi_{0} \: \scopa_0
\end{align*}

@@

The proofs of Theorem \ref{theorem:sparsityofdifferentialoperators} and Lemma \ref{lemma:sparsityofparametertransformationoperators} provide a way to calculate the non-zero entries of the operator matrices given in Definition \ref{def:differentialoperators} and Definition \ref{def:parametertransformationoperators}. We can simply use quadrature to calculate the 1D inner products, which has a complexity of $\bigO(N^3)$. This proves much cheaper computationally than using the 3D quadrature rule to calculate the 3D inner products, which has a complexity of $\bigO(N^4)$ \bstodoinline{correct the O(nd) complexity}.

@@

% FIGURES
\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.25]{poisson-u=wyexpx-N=60}
	\centering
	%\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.3]{solutionblocknorms-poisson-N=100}
        	%\label{fig:solutionblocknorms-poisson}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y,z) = - 2 e^x y z (2+x) + \genjacw^{(1,0)}(z) e^x (y^3 + z^2 y - 4xy - 2y)$. Right: The norms of each block of the computed solution of the Poisson equation with the given right hand side functions (here, $\bf{p}$ is simply a point on the spherical cap surface away from the pole and boundary). This demonstrates spectral convergence observed.}
	\centering
	\label{fig:poisson}
\end{figure}
%
%\begin{figure}[t]
%	\begin{subfigure}{0.3\textwidth}
%	\includegraphics[scale=0.3]{solution-exact-poisson-diskslice-alpha=0p2-beta=0p8-u=y3expx}
%	\centering
%	\end{subfigure}
%	\begin{subfigure}{0.3\textwidth}
%	\centering
%	\includegraphics[scale=0.3]{solution-exact-poisson-diskslice-alpha=0p2-beta=0p8-u=y3expx-actual}
%	\centering
%	\end{subfigure}
%	\begin{subfigure}{0.3\textwidth}
%	\includegraphics[scale=0.3]{solution-exact-poisson-diskslice-alpha=0p2-beta=0p8-u=y3expx-compare}
%	\centering
%	\end{subfigure}
%	\centering
%	\caption{The computed solution to $\Delta u = f$ with zero boundary conditions compared with the exact solution $u(x,y) = \Wiii (x,y) y^3 \exp(x)$. Left: Computed. Centre: Exact. Right: Plot of the error (colourbar is shown to demonstrate magnitude of the error is of the order $10^{-17}$)}
%	\centering
%	\label{fig:poissonexact}
%\end{figure}

\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.25]{helmholtz-f=wyexpx-N=60-n=300}
	\centering
	%\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.3]{solutionblocknorms-helmholtz-N=100}
        	%\label{fig:solutionblocknorms-poisson}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y,z) = y e^x \genjacw^{(1,0)}(z)$, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{(x_0^2 + z_0^2)}$ and $k = 100$. Right: The norms of each block of the computed solution of the Helmholtz equation with the given right hand side functions, with $k=20$ and the same function $v$ (again, $\bf{p}$ is simply a point on the spherical cap surface away from the pole and boundary). This demonstrates spectral convergence observed.}
	\centering
	\label{fig:helmholtz}
\end{figure}

\begin{figure}[t]
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.25]{biharmonic-f=erf-N=80-n=50}
	\centering
	%\label{fig:solution-biharmonic}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.3]{solutionblocknorms-biharmonic-N=100}
	\centering
        	%\label{fig:solutionblocknorms-biharmonic}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta^2 u = f$ with zero Dirichlet and Neumann boundary conditions with $f(x,y,z) = (1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))) \rho(z)^2$. Right: The norms of each block of the computed solution of the biharmonic equation with the given right hand side functions (again, $\bf{p}$ is simply a point on the spherical cap surface away from the pole and boundary). This demonstrates algebraic convergence observed.}
	\centering
	\label{fig:biharmonic}
\end{figure}
% END FIGURES
